---
title: "Exploratory Data Analysis: WNBA Draft Data"
author: "Hamza Mahamud"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(corrplot)
library(scales)
library(knitr)
library(kableExtra)
```

## 1. Load & Inspect Data

```{r load-data}
df <- read_csv("/Users/hamzamahamud/project-wnba-analysis/data/raw/wnbadraft.csv")

# Preview
glimpse(df)
head(df) |> kable() |> kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r summary-stats}
summary(df)
```

---

## 2. Data Cleaning

```{r clean-data}
# Separate players who played vs. those who never played
df_played <- df |> filter(!is.na(games) & games > 0)
df_no_play <- df |> filter(is.na(games) | games == 0)

cat("Players with WNBA stats:", nrow(df_played), "\n")
cat("Players with no recorded games:", nrow(df_no_play), "\n")
cat("% who never played:", round(nrow(df_no_play) / nrow(df) * 100, 1), "%\n")

# Classify player tiers by win shares
df_played <- df_played |>
  mutate(
    tier = case_when(
      win_shares >= 4    ~ "Elite",
      win_shares >= 1.5  ~ "Solid Contributor",
      win_shares >= 0    ~ "Fringe/Bench",
      TRUE               ~ "Negative Impact"
    ),
    tier = factor(tier, levels = c("Elite", "Solid Contributor", "Fringe/Bench", "Negative Impact")),
    pick_group = case_when(
      overall_pick <= 6  ~ "Top 6",
      overall_pick <= 12 ~ "Picks 7–12",
      overall_pick <= 24 ~ "Picks 13–24",
      TRUE               ~ "Picks 25+"
    ),
    pick_group = factor(pick_group, levels = c("Top 6", "Picks 7–12", "Picks 13–24", "Picks 25+"))
  )
```

---

## 3. Missing Data Overview

```{r missing-data}
df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "missing") |>
  mutate(pct_missing = round(missing / nrow(df) * 100, 1)) |>
  arrange(desc(missing)) |>
  kable(caption = "Missing values per column") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r never-played-bar}
# What share of each draft class never played?
df |>
  mutate(played = !is.na(games) & games > 0) |>
  group_by(year) |>
  summarise(
    total = n(),
    never_played = sum(!played),
    pct = never_played / total * 100
  ) |>
  ggplot(aes(x = factor(year), y = pct, fill = factor(year))) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct), "%")), vjust = -0.4, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "% of Draft Class That Never Played a WNBA Game",
    x = "Draft Year", y = "% of Picks"
  ) +
  theme_minimal(base_size = 13)
```

---

## 4. Draft Pick vs. Performance

### 4.1 Win Shares by Overall Pick

```{r ws-by-pick-scatter}
df_played |>
  ggplot(aes(x = overall_pick, y = win_shares, color = factor(year))) +
  geom_point(size = 3, alpha = 0.75) +
  geom_smooth(method = "loess", se = TRUE, color = "black", linewidth = 0.8, linetype = "dashed") +
  geom_text_repel(
    data = df_played |> filter(win_shares >= 4),
    aes(label = player), size = 3, max.overlaps = 10
  ) +
  scale_color_brewer(palette = "Set1", name = "Draft Year") +
  labs(
    title = "Win Shares vs. Draft Pick Number",
    subtitle = "LOESS curve shows overall trend; labeled players are top performers",
    x = "Overall Pick", y = "Career Win Shares (WNBA)"
  ) +
  theme_minimal(base_size = 13)
```

### 4.2 Points per Game by Pick Group

```{r ppg-pick-group-box}
df_played |>
  ggplot(aes(x = pick_group, y = points, fill = pick_group)) +
  geom_boxplot(alpha = 0.8, outlier.shape = 21, outlier.fill = "white") +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.5) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Points Per Game by Draft Pick Range",
    x = "Pick Group", y = "Points Per Game",
    fill = "Pick Group"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### 4.3 Win Shares per 40 Minutes by Pick Group

```{r ws40-pick-group}
df_played |>
  group_by(pick_group) |>
  summarise(
    mean_ws40 = mean(win_shares_40, na.rm = TRUE),
    median_ws40 = median(win_shares_40, na.rm = TRUE),
    n = n()
  ) |>
  ggplot(aes(x = pick_group, y = mean_ws40, fill = pick_group)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = round(mean_ws40, 3)), vjust = -0.4, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Average Win Shares per 40 Minutes by Pick Group",
    x = "Pick Group", y = "Mean WS/40"
  ) +
  theme_minimal(base_size = 13)
```

### 4.4 Linear Regression: Pick Number → Win Shares

```{r lm-pick-ws}
model_ws <- lm(win_shares ~ overall_pick, data = df_played)
summary(model_ws)
```

```{r lm-pick-ws-plot}
df_played |>
  ggplot(aes(x = overall_pick, y = win_shares)) +
  geom_point(alpha = 0.6, color = "#2c7bb6") +
  geom_smooth(method = "lm", color = "#d7191c", se = TRUE) +
  labs(
    title = "Linear Regression: Draft Pick → Win Shares",
    subtitle = paste0("Adj. R² = ", round(summary(model_ws)$adj.r.squared, 3),
                      "  |  p-value: ", format.pval(summary(model_ws)$coefficients[2, 4], digits = 3)),
    x = "Overall Pick", y = "Win Shares"
  ) +
  theme_minimal(base_size = 13)
```

### 4.5 Multiple Regression: Pick + Year → Win Shares

```{r lm-multi}
model_multi <- lm(win_shares ~ overall_pick + factor(year), data = df_played)
summary(model_multi)
```

### 4.6 Correlation Matrix (Performance Variables)

```{r corrplot}
perf_vars <- df_played |>
  select(overall_pick, games, win_shares, win_shares_40, minutes_played, points, total_rebounds, assists) |>
  drop_na()

cor_matrix <- cor(perf_vars)

corrplot(cor_matrix,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         number.cex = 0.75,
         tl.col = "black",
         tl.srt = 45,
         title = "Correlation Matrix: Draft & Performance Variables",
         mar = c(0, 0, 2, 0))
```

---

## 5. College → WNBA Success

### 5.1 Top Colleges by Draft Volume

```{r top-colleges-count}
df |>
  filter(!is.na(college) & college != "") |>
  count(college, sort = TRUE) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(college, n), y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "#c6dbef", high = "#08519c") +
  labs(
    title = "Top 15 Colleges by Number of WNBA Draft Picks",
    x = NULL, y = "Number of Draft Picks"
  ) +
  theme_minimal(base_size = 13)
```

### 5.2 Top Colleges by Average Win Shares (min. 3 draftees)

```{r top-colleges-ws}
df_played |>
  filter(!is.na(college) & college != "") |>
  group_by(college) |>
  summarise(
    mean_ws = mean(win_shares, na.rm = TRUE),
    n_players = n()
  ) |>
  filter(n_players >= 3) |>
  arrange(desc(mean_ws)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(college, mean_ws), y = mean_ws, fill = mean_ws)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0("n=", n_players)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "#fee8c8", high = "#e34a33") +
  labs(
    title = "Average Win Shares by College (min. 3 draftees with playing time)",
    x = NULL, y = "Mean Win Shares"
  ) +
  theme_minimal(base_size = 13)
```

### 5.3 College Win Shares Distribution (Top 8 Programs)

```{r college-ws-violin}
top_colleges <- df_played |>
  filter(!is.na(college) & college != "") |>
  count(college, sort = TRUE) |>
  slice_head(n = 8) |>
  pull(college)

df_played |>
  filter(college %in% top_colleges) |>
  ggplot(aes(x = reorder(college, win_shares, median), y = win_shares, fill = college)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Win Shares Distribution for Top 8 Most-Drafted Programs",
    x = NULL, y = "Win Shares"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### 5.4 Do International Players Perform Differently?

```{r intl-vs-domestic}
df_played <- df_played |>
  mutate(player_type = if_else(!is.na(former) & former != "", "International", "Domestic (College)"))

df_played |>
  group_by(player_type) |>
  summarise(
    n = n(),
    mean_ws = round(mean(win_shares, na.rm = TRUE), 2),
    mean_pts = round(mean(points, na.rm = TRUE), 2),
    mean_games = round(mean(games, na.rm = TRUE), 1)
  ) |>
  kable(caption = "Domestic vs. International Draftee Performance") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r intl-vs-domestic-plot}
df_played |>
  ggplot(aes(x = player_type, y = win_shares, fill = player_type)) +
  geom_boxplot(alpha = 0.8, outlier.shape = 21, outlier.fill = "white") +
  geom_jitter(width = 0.15, size = 2, alpha = 0.5) +
  scale_fill_manual(values = c("#66c2a5", "#fc8d62")) +
  labs(
    title = "Win Shares: Domestic (College) vs. International Players",
    x = NULL, y = "Win Shares"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

## 6. Year-over-Year Trends

### 6.1 Average Performance by Draft Year

```{r yearly-avg}
df_played |>
  group_by(year) |>
  summarise(
    n = n(),
    mean_ws = mean(win_shares, na.rm = TRUE),
    mean_pts = mean(points, na.rm = TRUE),
    mean_reb = mean(total_rebounds, na.rm = TRUE),
    mean_ast = mean(assists, na.rm = TRUE),
    mean_games = mean(games, na.rm = TRUE)
  ) |>
  kable(digits = 2, caption = "Mean Performance Stats by Draft Year") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r yearly-ws-bar}
df_played |>
  group_by(year) |>
  summarise(mean_ws = mean(win_shares, na.rm = TRUE)) |>
  ggplot(aes(x = factor(year), y = mean_ws, fill = factor(year))) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = round(mean_ws, 2)), vjust = -0.4, size = 4.5) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Average Win Shares by Draft Class Year",
    x = "Draft Year", y = "Mean Win Shares"
  ) +
  theme_minimal(base_size = 13)
```

### 6.2 Distribution of Win Shares Across Years

```{r yearly-ws-violin}
df_played |>
  ggplot(aes(x = factor(year), y = win_shares, fill = factor(year))) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.08, fill = "white", outlier.shape = NA) +
  scale_fill_brewer(palette = "Pastel2") +
  labs(
    title = "Win Shares Distribution by Draft Class",
    x = "Draft Year", y = "Win Shares"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### 6.3 Points, Rebounds & Assists Over Draft Years

```{r yearly-stats-facet}
df_played |>
  select(year, points, total_rebounds, assists) |>
  pivot_longer(cols = -year, names_to = "stat", values_to = "value") |>
  mutate(stat = recode(stat,
    "points" = "Points Per Game",
    "total_rebounds" = "Rebounds Per Game",
    "assists" = "Assists Per Game"
  )) |>
  group_by(year, stat) |>
  summarise(mean_val = mean(value, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = factor(year), y = mean_val, group = stat, color = stat)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  facet_wrap(~stat, scales = "free_y") +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Average Per-Game Stats by Draft Year",
    x = "Draft Year", y = "Average Value", color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### 6.4 Top Performers from Each Draft Class

```{r top-per-year}
df_played |>
  group_by(year) |>
  slice_max(win_shares, n = 5) |>
  select(year, overall_pick, player, college, games, win_shares, points, total_rebounds, assists) |>
  arrange(year, desc(win_shares)) |>
  kable(caption = "Top 5 Players by Win Shares in Each Draft Class") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## 7. Multivariate Regression: Predicting Win Shares

```{r lm-full}
model_full <- lm(win_shares ~ overall_pick + games + minutes_played + factor(year),
                 data = df_played)
summary(model_full)
```

```{r lm-residuals}
# Residual diagnostics
par(mfrow = c(2, 2))
plot(model_full)
par(mfrow = c(1, 1))
```

```{r lm-predicted-vs-actual}
df_played |>
  mutate(predicted_ws = predict(model_full, newdata = df_played)) |>
  ggplot(aes(x = predicted_ws, y = win_shares)) +
  geom_point(alpha = 0.6, color = "#2c7bb6") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_text_repel(
    data = df_played |>
      mutate(predicted_ws = predict(model_full, newdata = df_played),
             resid = win_shares - predicted_ws) |>
      filter(abs(resid) > 1.5),
    aes(label = player), size = 3
  ) +
  labs(
    title = "Predicted vs. Actual Win Shares",
    subtitle = "Red dashed line = perfect prediction; labeled = largest residuals",
    x = "Predicted Win Shares", y = "Actual Win Shares"
  ) +
  theme_minimal(base_size = 13)
```

---

## 8. Summary Findings

```{r summary-table}
tibble(
  Finding = c(
    "Draft pick predicts performance",
    "Top 6 picks outperform later picks",
    "Best college programs",
    "International vs. domestic",
    "Draft class quality by year",
    "Key predictors of win shares"
  ),
  Detail = c(
    "Earlier picks have higher average win shares, but relationship is noisy (check adj. R²)",
    "Top 6 picks average significantly more PPG and WS than picks 13+",
    "Programs like UConn, Baylor, Oregon, and South Carolina consistently produce high-WS players",
    "Domestic college players slightly outperform internationals on average, but sample is small",
    "Compare mean WS across years; later classes have fewer seasons to accumulate stats",
    "Games played and minutes are strongest predictors; pick number has modest independent effect"
  )
) |>
  kable(caption = "Key EDA Takeaways") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```
